## Are there unexpected patterns within the same variable over time?

**Reference Script**: comparability\\Comparability_Patterns_over_time.Rmd

**Description**: This analysis is only relevant for longitudinal and time-series analysis. In this case, the plausibility of a value in a certain time is examined in terms of the history of values for each variable or construct. Trends are evaluated in order to determine whether the values are extreme in comparison to values in other points in time. A discontinuity in the trend does not necessarily mean an error in the data, but it's useful to keep in mind its effects when conducting the analysis. Additionally, discrete jumps can also help to detect changes in the labels of categorical values. As a hypothetical example, in one year males were categorized using the value of 0 but in another a new value of 1 is used.   

We use tableplots (Tennekes et al., 2011) to help assess if the relationship between two or more variables exists. A tableplot can aid in the exploration of large data files by summarizing it over a set of variables. Each variable is represented in a column, where each row is an aggregate of a given number of units. For continuous variables, the length of the light blue bar represents the mean, and the dark blue bars represent the standard deviation. For discrete variables, different colors indicate the frequency of each category. The data is sorted by a relevant variable, thus allowing to assess the relationship with other variables.  

For this analysis, we will use tableplots sorted by time. Some of the advantages of using tableplots over simple two-way graphs is that it is possible to see if the discontinuity in one variables is associated with an event in other variables, and it's also possible to analyze changes in the standard deviation of the variables.  
  
  
### Patterns over time

By default, all key and domain variables will be included in the graph, unless the user specifies a different set of variables.

**Select variables:**
```{r patterns_over_time1}
##### SELECT VARIABLES #####
selected_var <- c(key_vars, domain_vars)
############################

# Check that selected variables are part of the key/domain/location/time variables. 
# If not, we select all key and domain variables
if (is.null(selected_var) | !all(is.element(selected_var, c(key_vars, domain_vars, location_vars, time_vars))) )  {
  selected_var <- c(key_vars, domain_vars)
}
```

**Tableplots:**
```{r patterns_over_time3}
# Select the first time variable 
time_var <- time_vars[1]

# Tableplot 
if (!is.null(time_var)) {
  
  #Convert to factor
  subsetted$time_factor <- factor(subsetted[[time_var]])
  time_var <- c("time_factor")
  
  options(ffmaxbytes = min(getOption("ffmaxbytes"),.Machine$integer.max * 12))
  
  selected_var2 <- c(selected_var, time_var)
  tableplot(subsetted, 
          select_string = selected_var2, 
          sortCol = time_var, scale = "lin", title = paste0("Tableplot sorted by: Time \nSubset: ",subset_text))
} else {
  print("No valid time variable")
}
```


**What to look for in tableplots**:  
- *Relation between variables*: Use the graph to check relationships between variables that are expected to happen. Also check if there are relationships that are apparent from the graph but that were not expected.  
- *Discontinuities in unsorted variables*: Among unsorted variables, look for discontinuities in continuous and categorical variables, and note the direction of the jump.
- *Range of values*: Look at the range of values of each variable to detect variables with particularly high or low range that may cast doubt about the validity of the values.  
- *High standard deviation or discrete jumps in the standard deviation*: Look for high standard deviations, which may indicate problems in the validity of the values. For example, in some cases missing values are codified with values such as 999999 or 0 which will increase the variability of the data. Also, look for drastic changes in the standard deviation.  
- *Highest and lowest values*: Check the highest and lowest values of the sorted variable and how they relate to the other variables.  
  
  
### Patterns over time by subgroup

In some cases, it is relevant to analyze the patterns over time for different subgroups. This is the case when we expect that data will exhibit different patterns in each subgroup. For example, in the case of welfare receipt, cases that only have children as recipients show certain characteristics that make them different from other cases, such as the exemption from federal time limits.

The user should select a categorical variable that identifies the different groups that want to be analyzed, which is recommended to have less than five categories. By default, the system will use the first domain variable.  

**Select subgroup variable:**
```{r patterns_over_time_bysubgroup1}
##### SELECT SUBGROUP VARIABLE #####
subgroup = domain_vars[1]
####################################

# If no subgroup variable is specified or if subgroup variable is not a key,
# domain, time, or location variable, or if there is more than one subgroup 
# variable, we skip this section.
if (is.null(subgroup) | 
    !all(is.element(subgroup, c(key_vars, domain_vars, location_vars, time_vars))) | 
    !(length(subgroup)==1) )  {
  print("Subgroup variable is not a key, domain, time, or location variable, or there is more than one subgroup variable")
  subgroup <- NULL
} else if (nlevels(as.factor(subsetted[[subgroup]])) >= 15) {
  print("Too many levels in the subgroup variable")
  subgroup <- NULL
}
```

**Tableplots:**
```{r patterns_over_time_bysubgroup2}

if (!is.null(subgroup) & !is.null(time_var)) {
  for ( i in levels(subsetted[[subgroup]]) ) {
    data5 <- subset(subsetted,subsetted[[subgroup]] == i)
    if (nrow(data5) > 0) {
      tableplot(data5, 
                select_string = selected_var2, 
                sortCol = time_var, scale = "lin", title = paste0("Tableplot sorted by: Time. Where ",subgroup, "= ", i,"\nSubset: ",subset_text))
    }
  }
} else {
  print("No valid subgroup or time variable")
}
```

**Interpretation**: In addition to the guidance provided in the previous subsection, check if there is any difference in the relationship patterns between variables across the different subgroups. Also check if data quality issues are concentrated in certain subgroups.  

