
## How are the data distributed?  Are there any values which seem atypical or implausible when compared to the overall distribution of the data?

**Reference Script**: outliers\\outliers.rmd

**Description**: It is always advisable to understand the distributional characteristics of the data that one is working with prior to investigating associative relationships and conducting other types of analysis.  Understanding which variables are qualitative, categorical, ordinal, and continuous, for example, allows the user to create appropriate specifications in models.  Additionally, knowing how the data are distributed gives the user a sense of the range of values that are likely for a given variable.  Sometimes this will suggest the application of some form of scaling to make the data more suitable for analysis.  It may also help uncover the presence of outliers which will require special handling.  Furthermore, sometimes variables are of limited interest when they only assume 1 or a few values since the lack of variation does not differentiate the records in meaningful ways.  The user is encouraged to discover other reasons for why the distributional characteristics of the data are important for their specific use case.

### Descriptive Statistics for Numeric Variables


```{r descriptives_num}

##### EDIT THIS SECTION TO DEFINE THE LOCATION TO STORE THE DESCRIPTIVE STATISTICS AND TO SELECT THE APPROPRIATE DATA FOR ANALYSIS#####

##### SELECT VARIABLES FOR DESCRIPTIVE STATISTICS OUTPUT#####
selected_var <- c(key_vars, domain_vars,id_vars,time_vars,location_vars)
############################
########################################################

# BASIC DATA QUALITY
num.file <- "dq_num.csv"
cat.file <- "dq_cat.csv"
checkDataQuality(data= analysis_file[,c(key_vars, domain_vars,id_vars,time_vars,location_vars)], out.file.num= num.file, out.file.cat= cat.file)

### Numeric Variables: Data Quality Check

numfile <- read.csv("dq_num.csv")
kable(numfile,caption="Numeric Variables descriptives")

```

`r cat("/n")`

**Interpretation**: Evaluate the descriptives statistics for each numeric variable and determine if the values and ranges are reasonable. Explore anomalous results in more detail.

### Descriptive Statistics for Character Variables


```{r descriptives_char}

### Character Variables: Data Quality Check

charfile <- read.csv("dq_cat.csv")
kable(charfile,caption="Character variables descriptives")

```

 
**Interpretation**: Evaluate the descriptives statistics for each character variable and determine if the unique values and frequency for each are appropriate. Explore anomalous results in more detail.

`r cat("/n")`
`r cat("/n")`

### Distributions, Extreme Values, and Outliers

For variables that are continuous it is appropriate and important to explore the distributional quality of the data.  The identification of skewness, deviation from normality, and outliers can be included in the list of important steps in the data evaluation process.  Traditional boxplots which use 4 quantiles (i.e. quartiles) can be helpful in understanding basic distributional characteristics (e.g. median, inter-quartile range) from certain types of data.  However, they are limited in their ability to differentiate between extreme values and outliers, and do not provide as granular distribution information as other methods especially in the tail of the distribution where it is often desirable to understand more.  
One method that satisfies these needs and can supplement the boxplot is the letter value plot.  The letter value plot dynamically determines the appropriate number of quantiles to use to best fit the data.  Use multiple analytical methods including boxplots, letter value plots, outlier detection and descriptive statistics to gain a more complete picture of the data and to inform the interpretation of analysis.

**Interpretation of Boxplots and Letter Value Plots**

Boxplots divide the ordered data into 4 groups such that the 25% of the data representing the lowest values are grouped into the first quartile.  The second quartile is composed of the next 25% lowest values and so on.  From this grouping the distribution is realized with the median represented by the middle bar that divides the two boxes.  The whiskers represent the range of the more extreme values of the data.  And the dots, if there are any, represent data that may be outliers.  Outliers are determined according to the method of "fences" as outlined by Tukey (1977).  In this method the inter-quartile range is multiplied by either 1.5 or 3 and the result is added or subtracted from the appropriate quartile.  Points that fall beyond these thresholds are identified as potential outliers.

The letter value plot approach divides the data into a determined number of quantiles as opposed to the automatic 4 in the boxplot.  The exact number of quantiles (boxes) that are created depends on the number of records, the distributional quality of the data, and the level of uncertainty in the quantiles' upper and lower bounds.  Similarly, the identification of outliers (dots) is performed in an adaptive way, flagging records that fall outside of the most extreme quantiles. For more details on letter value plots and the determination of outliers see Hofmann, Wickham, and Kafadar (2017).


```{r uni_outliers}

##### EDIT THIS SECTION TO DEFINE CONTINUOUS VARIABLES OF INTEREST#####
contvars <- as.character(numfile$X)

########################################################

testnums <- contvars
rm_vars <- as.character()
for (i in 1:length(testnums)) {
  min <- min(analysis_file[,testnums[i]],na.rm = T)
  max <- max(analysis_file[,testnums[i]],na.rm = T)
  if (min==max) print(paste0("Variable ", testnums[i], " assumes only 1 value across the entire dataset. That value is ",min,". Plots and outlier status will be suppressed."))
  if (min==max) rm_vars=rbind(rm_vars,testnums[i])
}
contvars=setdiff(contvars,rm_vars)

for (i in c(1:length(contvars))) {
  cat("\n")
  cat("\n","Distributions, Plots and Potential Outliers for",contvars[i])
  if (length(unique(analysis_file[[contvars[i]]])) < 30) {
    tablecont <-kable(table(analysis_file[[contvars[i]]],useNA="ifany",dnn=c(contvars[i])))
    print(tablecont)
  } else {
    print(histogram(analysis_file[[contvars[i]]],xlab=contvars[i]))
  }
  desc <- summary(analysis_file[[contvars[i]]])
  iqr <- desc["3rd Qu."] - desc["1st Qu."]
  
  cat("\n")
  print(kable(table(analysis_file[[contvars[i]]]> desc["3rd Qu."]+1.5*iqr,dnn="Mild High Outlier")))
  maxout<- analysis_file[which(analysis_file[[contvars[i]]]> desc["3rd Qu."]+1.5*iqr),]
  print(paste("Number of mild high outliers based on quartile method for variable",contvars[i],"is",nrow(maxout)))
  print(kable(table(analysis_file[[contvars[i]]]> desc["3rd Qu."]+3*iqr,dnn="Extreme High Outlier")))
  maxout<- analysis_file[which(analysis_file[[contvars[i]]]> desc["3rd Qu."]+3*iqr),]  
  print(paste("Number of extreme high outliers based on quartile method for variable",contvars[i],"is",nrow(maxout)))

  print(kable(table(analysis_file[[contvars[i]]]< desc["1st Qu."]-1.5*iqr,dnn="Mild Low Outlier")))
  minout <- analysis_file[which(analysis_file[[contvars[i]]]< desc["1st Qu."]-1.5*iqr),]
  print(paste("Number of mild low outliers based on quartile method for variable",contvars[i],"is",nrow(minout)))
  print(kable(table(analysis_file[[contvars[i]]]< desc["1st Qu."]-3*iqr,dnn="Extreme Low Outlier")))
  minout <- analysis_file[which(analysis_file[[contvars[i]]]< desc["1st Qu."]-3*iqr),]
  print(paste("Number of extreme low outliers based on quartile method for variable",contvars[i],"is",nrow(minout)))

  
  boxplot(analysis_file[[contvars[i]]],main=paste("Boxplot for",contvars[i],"variable"))
  
  LVboxplot(analysis_file[!is.na(analysis_file[[contvars[i]]]),contvars[i]], alpha = 0.95, k = NULL, perc = NULL,
            horizontal = TRUE, xlab = contvars[i], ylab = NULL, col = "grey30",
            bg = "grey90", width = 0.9, width.method = "linear",
            median.col = "grey10",
            main=paste("Letter Value Plot for",contvars[i],"variable"))
  
lv_table <- lvtable(analysis_file[!is.na(analysis_file[[contvars[i]]]),contvars[i]],k=determineDepth(nrow(analysis_file[!is.na(analysis_file[[contvars[i]]]),]),alpha=.05),alpha=.95)

lv_table<-cbind(lv_table,rep(0,nrow(lv_table)))
colnames(lv_table)[5]<-"lv_thresh_h"
lv_table<-cbind(lv_table,rep(0,nrow(lv_table)))
colnames(lv_table)[6]<-"lv_thresh_l"
lv_table=as.data.frame(lv_table)

med <- which(row.names(lv_table)=="M")

# HIGH OUTLIER
for (j in c((med+1):(nrow(lv_table)-1))) {
  lv_table[j,'lv_thresh_h']<-  ifelse(lv_table[j,'97.5%']>lv_table[j+1,'2.5%'] ,1,0)
}
# LOW OUTLIER
for (k in c((med-1):2)) {
  lv_table[k,'lv_thresh_l']<-  ifelse(lv_table[k,'2.5%']<lv_table[k-1,'97.5%'] ,1,0)
}

lv_table[is.na(lv_table[['lv_thresh_h']]),'lv_thresh_h'] <-  0
lv_table[is.na(lv_table[['lv_thresh_l']]),'lv_thresh_l'] <-  0

if (any(lv_table[['lv_thresh_h']]==1)) {
  lv_threshval_h<-min(subset(lv_table,lv_thresh_h==1,select="97.5%"))
  print(kable(table(analysis_file[[contvars[i]]]>=lv_threshval_h,dnn="Letter Value High Outliers")))
  print(paste0("Number of letter value high outliers for variable ",contvars[i]," is ",nrow(subset(analysis_file,get(contvars[i])>=lv_threshval_h)),".  Values above ",lv_threshval_h," are considered outliers with this method."))
}  else {
  print(paste("There are no letter value high outliers for variable",contvars[i]))
}

if (any(lv_table[['lv_thresh_l']]==1)) {
  lv_threshval_l<-max(subset(lv_table,lv_thresh_l==1,select="2.5%"))
  print(kable(table(analysis_file[[contvars[i]]]<=lv_threshval_l,dnn="Letter Value Low Outliers")))
  print(paste0("Number of letter value low outliers for variable ",contvars[i]," is ",nrow(subset(analysis_file,get(contvars[i])<=lv_threshval_l)),".  Values below ",lv_threshval_l," are considered outliers with this method."))
}  else {
  print(paste("There are no letter value low outliers for variable",contvars[i]))
}
}

```

**Interpretation**: The text output bring attention to variables which potentially outliers for various threshold settings.  It also identifies variables that assume only one value in across the entire dataset and therefore would be of little interest for analysis.  The boxplots and letter value plots help visualize the overall distribution while bringing attention to the potential outliers.

When outliers are identified it may indicate possible data quality issues.  It is suggested that the user research these data points in further detail to determine the accuracy of these values.  The inclusion of these data in analytical processes such as regression modeling may not necessarily be appropriate and could lead to incorrect conclusions.  The user is advised to consider these issues when performing data analysis.
